---
title: "Covid91 vaccine study Final2023"
author: "Cleo Ware"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="final.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))

male = filter(FinalData, sex =="M")
female = filter(FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
drugUser = filter(FinalData, DrugUser =="yes")

```
# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.


$H_0$ : There are NO statistically significant difference in the data, meaning the vaccine and placebo are equally as effective.

$H_A$ : There are statistically significant difference in the data, meaning the vaccine and placebo are NOT equally as effective.

# Methods

When looking at the data, this seems like its going to be categorical explained by another categorical. We will create a separate data set for each sub population. Then use the helper file CandP to create a bar chart and display the data. Then we will use xtabs for percentages on the data.Given that we are only testing two types, we will not need to use any loops to measure multiple variables. 


# Results


## Males

## Graphical Results

```{r}
barchartGC(~infected + treatment,data=male)
barchartGC(~infected + treatment,data=male, type="percent")
```

The barchart above is comparing the graphical results for the Males group data set. It compares the rates of contracting Covid 19 with vaccination status. With the left side showing people who have contracted Covid 19, meaning they tested positive, while the right side shows males who have not contracted Covid 19.

Based on the data from the barchart, we can see that more than twice the amount of people taking the placebo contracted Covid 19. While those in the group who had taken the vaccine had a much lower number. Indicating the vaccine does have an effect.   


### Numerical Results

```{r}
tableMale <- xtabs(~infected + treatment, data=male)
rowPerc(tableMale)
colPerc(tableMale)
```


The numerical results above confirm what the barchart data is indicating. Which is that people who have taken the vaccicne tend to contract Covid 19 half as much as the group that has taken only the placebo. The placebo group contracted Covid 19 at approximately 70% of the time. While the group with the vaccine contracted Covid 19 30% of the time. Based on the results, the vaccine seems to be effective on males in this group. 



### Inferential Results

```{r}
chisq.test(tableMale)
chisqtestGC(tableMale)
maleFisher <- fisher.test(tableMale)
maleFisher
```


The Chi Squared test shows that the P value for the tested data is much smaller than the 0.05 limit. The results show $1.098\times10^{-13}$. This very small number shows that there is a difference in the data. Also, the Chi Squared statistic is 55, which is larger than the degree of freedom. Thus, the odds of an unvaccinated man getting Covid 19 is 2.2 times greater than a man with the vaccine.   



### Confidence Interval

```{r}
oddsMale = maleFisher$estimate/(maleFisher$estimate+1)*100
print(oddsMale)
maleLow=(maleFisher$conf.int[1]/(maleFisher$conf.int[1]+1))*100
maleHigh=(maleFisher$conf.int[2]/(maleFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Males: ",maleLow," to ",maleHigh)
```

We can see from the results that the confidence interval is >50% with a normal vaccine threshold of >60%. This is enough data to confidently say this vaccine is effective at preventing Covid 19 for this group. 

## Females


### Graphical Results

```{r}
barchartGC(~infected + treatment,data=female)
barchartGC(~infected + treatment,data=female, type="percent")
```

The above barchart compares the graphical results of only the Female group data collected. The comparison shows rates of contracting Covid 19 compared to whether or not the group has the vaccine. The left side shows females who contracted Covid 19 while the right shows females that have not. The barchart also shows that those females that took a placebo are a third more likely contracted Covid 19 than females who have taken the vaccine. 


### Numerical Results

```{r}
tableFemale <- xtabs(~infected + treatment, data=female)
rowPerc(tableFemale)
colPerc(tableFemale)
```


The numerical results above confirms the suggestion of the graphical data that those who took the placebo contracted Covid 19 approximately 60% of the time. While those females who are have taken the vaccine are contracted Covid 19 40% of the time. Thus showing that the vaccine is generally effective. However, it is slightly less effective than the male groups.

### Inferential Results

```{r}
chisq.test(tableFemale)
chisqtestGC(tableFemale)
femaleFisher <- fisher.test(tableFemale)
femaleFisher
```

The Chi Squared test shows that the P value for the data is smaller than the 0.05 limit. With the number being $2.732\times10^{-4}$.Showing that there is statistical significance.The Chi Squared statistic 13, which is also higher than the degrees of freedom for this group's results. Thus, the odds of an unvaccinated female contracting Covid 19 is 1.5 times higher than a female that has taken the vaccine contracting Covid 19.   


# Confidence Interval

```{r}
oddsFemale = femaleFisher$estimate/(femaleFisher$estimate +1)*100
print(oddsFemale)
femaleLow=(femaleFisher$conf.int[1]/(femaleFisher$conf.int[1]+1))*100
femaleHigh=(femaleFisher$conf.int[2]/(femaleFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Females: ",femaleLow," to ",femaleHigh)
```

As we can see from the results, we again see that the confidence interval is >50% and includes the 60% threshold. Showing that the vaccine is effective. However, its less effective than we saw in the male group results.


## LGBTQ

### Graphical Results

```{r}
barchartGC(~infected + treatment,data=lgbtq)
barchartGC(~infected + treatment,data=lgbtq, type="percent")
```

The above barchart is comparing the graphical results of placebo and vaccine data for only people from the LGBTQ group.Comparing whether or not the vaccine had any effect on the odds of catching Covid 19. The left side of the barchart shows people who contracted Covid 19 while taking the placebo. While the right side shows the people who have taken the vaccine and contracted Covid 19. As we see on the barchart, people who have taken the vaccine appear to be 5 times more likely to contract Covid 19 than those who took a placebo. 



### Numerical Results

```{r}
tableLGBTQ <- xtabs(~infected + treatment, data=lgbtq)
rowPerc(tableLGBTQ)
colPerc(tableLGBTQ)
```



The numerical results compound on what the graphical results were telling us, the placebo group contracted Covid approximately 15% of the time where the vaccinated group contracted Covid 85% of the time. We can see that the vaccine is not effective in this testing scenario, in fact it is inversely effective and seems to put this group at a higher risk of contracting Covid 19.



### Inferential Results

```{r}
chisq.test(tableLGBTQ)
chisqtestGC(tableLGBTQ)
lgbtqFisher <- fisher.test(tableLGBTQ)
lgbtqFisher
```


The p-value of $6.668\times10^{-8}$ is much smaller than the suggested 0.05 threshold. Showing that this reverse effect of the vaccine is indeed real. The Chi-Squared statistic, 29, supports this as it is larger than the degrees of freedom. The odds for an unvaccinated LGBTQ person getting Covid-19 compared to a vaccinated one are about 1:5. 



### Confidence Interval

```{r}
oddsLGBTQ = lgbtqFisher$estimate/(lgbtqFisher$estimate +1)*100
print(oddsLGBTQ)
lgbtqLow=(lgbtqFisher$conf.int[1]/(lgbtqFisher$conf.int[1]+1))*100
lgbtqHigh=(lgbtqFisher$conf.int[2]/(lgbtqFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for LGBTQ: ",lgbtqLow," to ",lgbtqHigh)
```

We see from the results, the confidence interval is 7-26%. This is far lower than 50%. Thus, we can say that this is not a very good vaccine.


## Druggies


### Graphical Results

```{r}
barchartGC(~infected + treatment,data=drugUser)
barchartGC(~infected + treatment,data=drugUser, type="percent")
```


The barchart above shows the graphical results for all people identified on drugs or druggies and how they contract Covid 19 based on whether or not they have the vaccine. The left shows positive right shows negative. In this instance it was five or six times more likely that a person who had contracted Covid 19 had also received the vaccine and not the placebo. This could suggest that drug users are already much more likely to contract Covid 19 because of Compromised immune systems. There's also the chance of a reverse affect from the vaccination.Which would require a separate study. 


### Numerical Results

```{r}
tableDrug <- xtabs(~infected + treatment, data=drugUser)
rowPerc(tableDrug)
colPerc(tableDrug)
```



The numerical results concur with what the graphical results show. The placebo group contracted Covid 19 approximately 10% of the time while the vaccinated group contracted Covid 19 90% of the time. We can see that the vaccine is not effective in this group. In fact, it is inversely effective and seems to put this group at a higher risk of contracting Covid 19. This group suffers from even more ill effects than the already negatively effected LGBTQ group.



### Inferential Results

```{r}
chisq.test(tableDrug)
chisqtestGC(tableDrug)
drugFisher <- fisher.test(tableDrug)
drugFisher
```


The Chi Squared test shows that the P value for the tested data is much smaller than the lower limit point of 0.05. The number is $2.842\times10^{-9}$. This small number indicates that there is a statistical difference in the data. The Chi Squared statistic is also 35, being larger than our degrees of freedom of one from this data. The odds of an unvaccinated person contracting Covid 19 in this group are about 1:8.



### Confidence Interval

```{r}
oddsDrug = drugFisher$estimate/(drugFisher$estimate +1)*100
print(oddsDrug)
drugLow=(drugFisher$conf.int[1]/(drugFisher$conf.int[1]+1))*100
drugHigh=(drugFisher$conf.int[2]/(drugFisher$conf.int[2]+1))*100
cat("\n","Condifence interval for Drug Users: ",drugLow," to ",drugHigh)
```

We see that the confidence interval is 4-21% far lower than 50% so we can say that this is also not a very effective vaccine.


# Overall Results and Conclusions

We see that for all categories we can reject the Null Hyposthesis but for different reasons, some groups had good results where the vaccine was effective and put people at a lower risk of contracting Covid 19 while others appareled to be at an even higher risk if they received the vaccine. We cannot tell from the data the exact reason for the discrepancy but it would appear that the male and female groups received benefits from the vaccine and are in a lower risk.While on the other hand, LGBTQ and Drug Users have very negative results from the vaccine. 

## Males

With the male group, we can see from the data that they had the best results from the vaccine. The results from percentage charts and bar charts show that a males were less likely to contract Covid 19 if they got the vaccine. We also saw that the difference in statistically significant were small P-Values and high Chi-Squared statistics.Therefore, we can reject the Null Hypothesis in favor of the alternative and determine that the vaccine IS effective in this group.


## Females

With the female group, we can see from the data that females had positive results from the vaccine. The results from percentage charts and bar charts show that a females were less likely to contract Covid 19 if they got the vaccine. We also saw that the difference in statistically significant was small P-Values and high Chi-Squared statistics. Therefore, we can reject the Null Hypothesis in favor of the alternative and determine that the vaccine is indeed effective in this group. Although, it was less effective than the male group.



## LGBTQ

With the LGBTQ group, we can see from the data that this group had negative results from the vaccine. The results from percentage charts and bar charts showed that LGBTQ people were more likely to contract Covid 19 if they got the vaccine. We also saw that the difference in statistically significant was small P-Values and high Chi-Squared statistics. Therefore, we can reject the Null Hypothesis in favor of the alternative and determine that the vaccine was not effective in this group. In fact, it appeared to put this group at a higher risk.


## Drug Users

With the drug users group, we can see from the data that they had the worse results from the vaccine. The results from percentage charts and bar charts show that drug users are more likely to contract Covid 19 if they got the vaccine. We can also see that the difference in statistically significant with small P-Values and high Chi-Squared statistics. Therefore, we can reject the Null Hypothesis in favor of the alternative and determine that the vaccine was not effective in this group. It also appears to put this group at a higher risk. More so than the LGBTQ group.
